---
title: "A Statistical Analysis of Global Temperature Trends"
subtitle: "Project - Team 10-7: The Outliers"
author: "By Morgan Pruchniewski, Scott Burstein, Katie Zhou"
date: "11/22/2020 "
output: 
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse) 
library(lubridate)
library(maps)
library(sp)
library(RColorBrewer)
library(paletteer)
library(ggthemes)
#install.packages('rworldmap', dependencies=TRUE)
# from https://cran.r-project.org/web/packages/rworldmap/vignettes/rworldmap.pdf
library(rworldmap)
```

# Introduction

One of the biggest environmental issues facing our planet has been climate
change. As fossil fuel emissions have increased, greenhouse gases have begun
trapping heat in our atmosphere that has created severe impacts. These include
increased global temperature and rising ocean levels, in addition to increased
natural disasters. [NASA.gov Source](https://climate.nasa.gov/effects/). 
[USGS.gov Source](https://www.usgs.gov/faqs/how-can-climate-change-affect-natural-disasters-1?qt-news_science_products=0#qt-news_science_products)

Climate change is a significant issue that has also become a hot-button 
political topic in recent years because of the failure of certain individuals, 
political parties, and corporations to take action: some people do in fact deny 
the existence of climate change/global warming despite a strong body of 
scientific support. 
[Basis of climate change denial Source](https://www.sciencealert.com/the-five-corrupt-pillars-of-climate-change-denial) 
The main argument behind this is that the Earth is always 
changing, and that the recent increase in global land and ocean temperatures
is not significant enough to support global warming.

In this report, we will be analyzing a dataset containing recorded land and 
ocean temperatures over the last 100+ years to examine whether there is evidence 
to suggest a statistically significant increase in land temperature over the 
last 30 years. 

## Research Questions
Our main research question will be: 

### Is there significant evidence to support the existence of climate change?  

To evaluate this, we will conduct comprehensive hypothesis tests on four 
sub-questions.

*1. Is there evidence to suggest a statistically significant increase in mean
earth surface temperature from early-20th-century levels to what the data show 
for more recent years?*. 

For this question, we will use $\alpha = 0.01$.

Our hypotheses for this question are as follows:

$H_o:$
The mean anomaly of annual temperature averages with respect to the average 
temperature before 1980 is not greater than 0 (the mean value from before 1980 
is used as the reference value for calculating anomalies, so the mean anomaly of
pre-1980 values is assumed to be 0).  

$H_a:$
The mean anomaly of annual temperature averages with respect to the average 
temperature before 1980 is greater than 0 (the mean anomaly of data before 1980 
with respect to its average temperature).

We predict that there will be sufficient evidence to support a statistically 
significant increase as detailed above, which brings us to the next question:  

*2. Is the earth changing/increasing temperature at a faster rate now than it 
was in the early 20th century?*

For this question, we will record the observed average rates of change (slope) 
for every five-year period, and test whether the mean slope after 1980 is 
greater than the mean slope before 1980 by a statistically significant margin. 
Because there are only 23 five-year segments between 1900 and 2010, we will 
use a lower $\alpha = 0.01$ level of 0.05.

Our hypotheses for this question are as follows:

$H_o:$
The rate at which the average global land temperature increased after 
1980 is not greater than the rate of temperature increase starting from 1900
based on their respective linear models.

$H_a:$
The rate at which the average global land temperature increased after 
1980 is greater than the rate of temperature increase starting from 1900
based on their respective linear models.  

Relating to the previous question, which looks at the rates of presumed 
increase, we will also evaluate net change (positive of negative).  

*3. Does the data provide evidence of a greater degree of net annual temperature
fluctuation after 1980 than before 1980?*

Finally, we will evaluate how the changes in earth surface temperature may 
differ geographically. With Europe and North America having had large roles in 
industrialization, globalization, and technological developments, all of which 
have been cited as major contributors to climate change, we wanted to consider 
if this might be manifested in the amount of climate change exhibited in the 
last century in those continents.

*4. Has North America experienced a greater change in annual average 
temperatures from the first half of the 20th century to 2010 than other 
continents? Has Europe?*

Our hypotheses for this two-part question are as follows:

**North America**  

$H_{o}:$
The change in mean temperature from before 1950 to post-2010 in North 
America is equal to the change in temperature observed in the rest of the world.  
$H_{a}:$
The change in mean temperature from before 1950 to post-2010 in North 
America is not equal to the change in temperature observed in the rest of the 
world.  

**Europe**  

$H_{o}:$
The change in mean temperature from before 1950 to post-2010 in Europe
is equal to the change in temperature observed in the rest of the world.  

$H_{a}:$
The change in mean temperature from before 1950 to post-2010 in Europe
is not equal to the change in temperature observed in the rest of the world.

## Data Sources:

We found this dataset on Kaggle 
[climate-change-earth-surface-temperature-data](https://www.kaggle.com/berkeleyearth/climate-change-earth-surface-temperature-data).

The data was originally compiled by the Berkeley Earth Data Lab from 
16 pre-existing archives, and it was updated to Kaggle in 2017. It is important
to note that many of the earlier years include NA values, which is why we 
focused on analysis on the 1900s.

## Types of Information Present in Data:

### By Time and Location (for `GlobalLandTemperaturesByMajorCity`)

Each observation in `GlobalLandTemperaturesByMajorCity` is a city and its 
respective land temperature, coordinates, and Country, which will be used to 
investigate changes in climate over time. There are 239,177 rows in the major 
city dataset. 

### By Time (for `GlobalTemperatures`)

Each observation in `GlobalTemperatures` is a numeric date/time value 
(described below) and its respective land temperature, max temperature, 
min temperature, ocean temperature, and relevant uncertainties to said 
variables, which will be used to investigate changes in climate over time. 
There are 3,192 rows in the global dataset. 

### Relevant Variables:

The relevant variables in the `major_city` dataset include dt (date and time of 
the recorded temperature, a discrete numeric), AverageTemperature (the average 
temperature for each city, a continuous numeric), city (nominal categorical), 
country (nominal categorical), latitude (continuous numeric), and longitude 
(continuous numeric).

The relevant variables in the `global` dataset include dt (date and time of the 
recorded temperature, a discrete numeric), LandAverageTemperature (the average 
global land temperature at that time, continuous numeric), LandMaxTemperature 
(the highest recorded land temperature of that year, a continuous numeric), 
LandMinTemperature (the lowest recorded land temperature of that year, a 
continous numeric), and LandAndOceanAverageTemperature (the average of land and 
ocean temperature averages for that year, a continuous numeric).

```{r read-in-data, message = FALSE, echo = FALSE}
#read in datasets
major_city <- read_csv("data/GlobalLandTemperaturesByMajorCity.csv")
global <- read_csv("data/GlobalTemperatures.csv",
                    col_types = "Tdddddddd")
df <- left_join(major_city, global) 
```

```{r data-processing, echo = FALSE}
df$Latitude<-as.numeric(gsub("N|E|S|W", "",df$Latitude))*
  ifelse(grepl("S",df$Latitude),-1,1)

df$Longitude<-as.numeric(gsub("N|E|S|W", "",df$Longitude))*
  ifelse(grepl("W",df$Longitude),-1,1)

df <- df %>%
  mutate(month = month(dt),
         year = year(dt),
         season = case_when(
           month < 3 | month == 12 ~ "winter",
           month >= 3 & month < 6 ~ "spring",
           month >=6 & month < 9 ~ "summer",
           month >= 9 & month < 12 ~ "autumn"
         )) 

df$Country <- replace(df$Country, df$Country == 
                        "Congo (Democratic Republic Of The)", "Congo")
```

# Methodology

## Initial Visualizations:

```{r ggplot-theme, echo = FALSE}
theme_economist_mod <- function() {
  theme_economist() + 
  theme(
    rect = element_rect(fill = "#E5E9F0", colour = NA,
                              linetype = 1),
    panel.background = element_rect(linetype = 0),
    plot.background = element_rect(fill = "#E5E9F0", colour = NA),
    line = element_line(colour = "white"),
    axis.line = element_line(size = rel(1.5)),
    axis.title.y = element_text(margin = unit(c(0, 2, 0, 0), "mm")),
    axis.title.x = element_text(margin = unit(c(4, 0, 0, 0), "mm")))
}

seasoncolors <- c("summer" = "#D08770","autumn" = "#EBCB8B",
             "winter" = "#88C0D0","spring" = "#B48EAD")
gradientcolors <- c(
  "#B48EAD","#5E81AC","#A3BE8C","#EBCB8B","#D08770","#BF616A")
```

```{r visualization-by-season, message = FALSE, echo = FALSE}
ggplot(data=df, aes(x=dt, y=LandAverageTemperature)) +
  geom_point(aes(color=season), size=0.8) +
  labs(title = "Global average land temperatures since 1849",
       x = "Year",
       y = "Average land temp.",
       color = "Season") +
  scale_color_manual(values=seasoncolors) +
  theme_economist_mod() +
  guides(col = guide_legend(override.aes = list(size=4)))
```

Based on this initial visualization, there seems to be a high amount of 
uncertainty in the average land temperature in the earlier data. Going forward, 
we will only use the data after 1900. 
Overall, across all seasons, there seems to be an increasing trend in average 
land temperature after roughly 1975. 

```{r filter-years, echo = FALSE}
df <- df %>%
  filter(dt >= "1900-01-01")
```

```{r temp-oceans, echo = FALSE}
ggplot(data=df, aes(x=dt, y=LandAndOceanAverageTemperature)) +
  geom_point(aes(color=season), size=0.8) + 
  labs(title = "Global average land + ocean temperatures since 1900",
       x = "Year",
       y = "Average land + ocean temp.",
       color="Season") + 
  scale_color_manual(values=seasoncolors) +
  theme_economist_mod() +
  guides(col = guide_legend(override.aes = list(size=4)))
```

Based on the second visualization, the average land/ocean temperature seems to 
increase more steeply after 1975 than the average land temperature. The values 
for spring and autumn in the land/ocean data are also much more similar than 
they are for just land.

```{r annual-avg-viz, message = FALSE, echo = FALSE}
annual_avg <- df %>%
  group_by(year) %>%
  summarize(avg = mean(LandAverageTemperature),
            max_avg = mean(LandMaxTemperature),
            min_avg = mean(LandMinTemperature)) 

ggplot(data=annual_avg, aes(x=year, y=avg)) +
  geom_point(aes(color=avg), size=3) + 
  labs(title = "Global average land temperatures since 1900",
       x = "Year",
       y = "Average land temp.") +
  theme_economist_mod() +
  scale_colour_gradientn(colours = gradientcolors) + 
  theme(legend.position="right")
```

When we plot the average annual land temperature (an average of the data points 
from each month), we can visualize a more extreme change in average land 
temperature over time than when each month is plotted individually. This is 
marked by a more severe uptick since 1980.

```{r facet-season, echo = FALSE}
ggplot(data=df, aes(x=dt, y=LandAverageTemperature)) +
  geom_point(aes(color=LandAverageTemperature), size=.3) + 
  facet_wrap(~season) +
  labs(title="Global average land temperatures",
       subtitle="faceted by season",
       y="Average land temp.",
       x = "Year",
       color = "Average \nland temp.") +
  theme_economist_mod() +
  scale_colour_gradientn(colours = gradientcolors) +
  theme(legend.position="right")
```

```{r highs-lows-visualization, echo = FALSE}
ggplot(data=annual_avg, aes(x = year)) +
  geom_line(aes(y = max_avg), color="#5E81AC", size=0.3) +
  geom_line(aes(y = min_avg), color="#5E81AC", size=0.3) + 
  geom_point(aes(y = max_avg), size=0.6, color="#434C5E") +
  geom_point(aes(y = min_avg), size=0.6, color="#434C5E") +
  labs(title = "Annual global average highs and lows",
       x = "Year",
       y = "Average global extremes") +
  theme_economist_mod() +
  theme(
    axis.title.y = element_text(margin = unit(c(0, 2, 0, 0), "mm"))
  )
```
  
The upward trend visualized in the first few graphs is consistent over each 
season, as well as for the annual average temperature highs and lows.

```{r country-avg-changes, echo = FALSE, message=FALSE}
bycountrypre1950 <- df %>%
  filter(!is.na(AverageTemperature),
         year < 1950) %>%
  group_by(Country) %>%
  summarize(avg = mean(AverageTemperature),
            lat = mean(Latitude),
            long = mean(Longitude)) 
  

bycountry2010s <- df %>%
  filter(!is.na(AverageTemperature),
         year > 2009) %>%
  group_by(Country) %>%
  summarize(avg = mean(AverageTemperature),
            lat = mean(Latitude),
            long = mean(Longitude))

changesbycountry <- tibble(country=bycountry2010s$Country, 
                           change = bycountry2010s$avg - bycountrypre1950$avg,
                           lat = bycountry2010s$lat,
                           long = bycountry2010s$long)
```

```{r visualization-by-country, echo = FALSE}
legend_mod <- function() {
  theme(legend.position="right",
        legend.title = element_text(size=10),
        legend.title.align = 1,
        legend.text = element_text(size=8),
        legend.box.spacing = unit(0.05, "cm"),
        legend.margin = margin(0,0,0,0, "cm"),
        legend.key.size = unit(0.5,"cm")) 
}

ggplot(data=changesbycountry,aes(x=reorder(country,-change),y=change)) +
  geom_bar(stat="identity", aes(fill=change)) + 
  labs(
    title="Average change from pre-1950 data to post-2010 data",
    subtitle = "by country",
    x="Country",
    y = "Change in average temperature",
    fill = "Change in \naverage \ntemp.") + 
  theme_economist_mod() +
  theme(axis.title.y = element_text(margin = unit(c(0, 1, 0, 0), "mm")),
    axis.text.x = element_text(angle = 90, size = 8,hjust=1),
    plot.subtitle = element_text(size = 17, hjust = 0.07, vjust = 0.5)) +
  scale_fill_gradientn(colours = gradientcolors) +
  legend_mod() +
  guides(fill = guide_colorbar(label.position = "left"),label.hjust=1) 
```

In this visualization we can see the countries that have had the greatest change
in average temperature from before 1950 to 2010-2013.

# Results

## Question 1

***Is there evidence to suggest a statistically significant increase in mean
earth surface temperature from early-20th-century levels to what the data show 
for more recent years?***

### Null Hypothesis:

$H_o: \mu_{post1980} \leq \mu_{pre1980}$

The mean anomaly of annual temperature averages with respect to the average 
temperature before 1980 is not greater than 0 (the mean value from before 1980 
is used as the reference value for calculating anomalies, so the mean anomaly of
pre-1980 values is assumed to be 0). 

### Alternative Hypothesis:

$H_a: \mu_{post1980} > \mu_{pre1980}$ 
 
The mean anomaly of annual temperature averages with respect to the average 
temperature before 1980 is greater than 0 (the mean anomaly of data before 1980 
with respect to its average temperature).

```{r per-year-diffs-loop, echo = FALSE}
differences <- c(nrow(annual_avg))

for (i in 1:nrow(annual_avg) - 1) {
  differences[i] <- annual_avg$avg[i+1] - annual_avg$avg[i]
}
```

```{r per-year-diffs-create-df, echo = FALSE}
differences <- tibble(differences)
differences <- tibble::rowid_to_column(differences, "year")
differences <- differences %>%
  mutate(year = 1900 + year)
```

```{r per-year-diffs-viz, echo = FALSE}
ggplot(data=differences, aes(x=year, y=differences)) +
  geom_bar(aes(fill=differences), stat="identity") +
  labs(title = 
         "Year-to year differences in annual \naverage temperatures since 1900",
       fill = "Differences",
       x = "Year",
       y = "Change in annual avg. from previous year") + 
  theme_economist_mod() + 
  scale_fill_gradientn(colours = c("#EBCB8B","#D08770", "#BF616A")) +
  legend_mod()
```

Based on this visualization, someone could potentially try to counter the 
argument for the existence of climate change, because there is a large 
fluctuation of positive and negative changes from year to year. Therefore, we 
then created a new data-frame to measure the overall difference between each 
year's average temperature and the average temperature across the 20th century, 
as shown below, and are using the comparison to the 20th century average for our 
hypothesis test.

```{r create-anomalies-data, echo = FALSE}
avg_to1980 <- annual_avg %>%
  filter(year < 1980) %>%
  summarize(mean(avg)) %>%
  pull()

annual_avg <- annual_avg %>%
  mutate(anomaly = avg - avg_to1980,
         post1980 = year >= 1980)
```

```{r visualization-anomalies, echo = FALSE}
ggplot(data=annual_avg, aes(x=year, y=anomaly)) +
  geom_bar(stat="identity", aes(fill=anomaly)) +
  labs(title=
  "Anomalies of average temperatures \nwith respect to 1900-1980 average",
    x = "Year",
    y = "Difference from 1900-1980 average",
    fill = "Difference amount") + 
  theme_economist_mod() + 
  scale_fill_gradientn(colours = gradientcolors) +
  legend_mod()
```

```{r anomaly-bootstrap, echo = FALSE}
post <- annual_avg %>%
  filter(post1980 == TRUE) 

pre <- annual_avg %>%
  filter(post1980 == FALSE)

set.seed(7)

boot_dist <- numeric(5000)

for (i in 1:5000) {
  post_mean <- post %>%
    slice(sample(1:nrow(post), replace = TRUE)) %>%
    summarize(mean(anomaly)) %>%
    pull()
  pre_mean <- pre %>%
    slice(sample(1:nrow(pre), replace = TRUE)) %>%
    summarize(mean(anomaly)) %>%
    pull()
  boot_dist[i] <- post_mean - pre_mean
}

boot_dist <- tibble(boot_dist)
```

```{r anomaly-ci, echo = FALSE}
diff_bootstrap_interval <- boot_dist %>%
  summarize(lower = quantile(boot_dist, 0.005),
            upper = quantile(boot_dist, 0.995)
            )
diff_bootstrap_interval
```

Our 99% confidence interval for this hypothesis test is (0.5514, 0.8941).
We can be 99% confident that the true mean anomaly of annual temperature 
averages after 1980 with respect to the average temperature before 1980 is 
between these bounds.

```{r anomaly-testing, echo = FALSE}
pre_mean <- pre %>%
  summarize(mean(anomaly)) %>%
  pull()

post_mean <- post %>%
  summarize(mean(anomaly)) %>%
  pull()

obs_diff <- post_mean - pre_mean

boot_mean <- boot_dist %>%
  summarize(mean(boot_dist)) %>%
  pull()

shifted <- boot_dist %>%
  mutate(shifted = boot_dist - boot_mean)

shifted %>%
  mutate(extreme = if_else(shifted >= obs_diff, 1, 0)) %>%
  summarize(p_val = mean(extreme)) %>%
  pull()
```

```{r compare-to-20th-sim-viz, echo = FALSE}
ggplot(data = shifted, aes(x = shifted)) +
  geom_histogram(color = "#4C566A", fill = "#88C0D0",bins=50) +
  geom_vline(xintercept = obs_diff, color = "#BF616A", lwd=1) +
  annotate("text", 
           x = obs_diff - 0.045, y = 350, 
           label= paste("Observed difference: \n", round(obs_diff, 4)),
           angle = 90, size = 5) + 
  labs(title = "Simulated null distribution of post-1980 anomalies",
       subtitle = 
         "Average annual land temp. after 1980 minus average temp. before 1980",
       x = "Mean difference in temperature",
       y = "Count") +
  theme_economist_mod() 
```

```{r compare-to-20th-clt, echo = FALSE}
t.test(anomaly ~ post1980,
  data = annual_avg,
  mu = 0,
  var.equal = FALSE,
  alternative = "less",
  conf.level = 0.99)
```

In both approaches used in this analysis, the p-value is well under 
$\alpha = 0.01$, so we reject this null hypothesis.

The p-value in this approach is the probability of obtaining results in which 
the post-1980 data exhibit an anomaly with respect to the pre-1980 average 
temperature as large as observed or greater, assuming under the null hypothesis 
that the post-1980 temperatures are not significantly different from the 
pre-1980 temperatures. From both tests, the p-value was extremely small: 
we observed a 0 p-value in the simulation-based approach. 

In context of our question, this indicates that there is evidence to support a 
statistically significant increase in temperature over the 20th century. This is
important because this is one of the main arguments used against climate change.


## Question 2

***Is the earth changing/increasing temperature at a faster rate now than it was 
in the early 20th century?***

### Null Hypothesis:

$H_o: \beta_{post1980} \leq \beta_{pre1980}$ 

The rate at which the average global land temperature increased after 1980 
is not greater than the rate of temperature increase starting from 1900
based on their respective linear models.

### Alternative Hypothesis:

$H_a: \beta_{post1980} > \beta_{pre1980}$ 

The rate at which the average global land temperature increased after 1980 
is greater than the rate of temperature increase starting from 1900
based on their respective linear models.

```{r time-period-lm, echo = FALSE}
m_main <- lm(avg ~ year, data=annual_avg)
tidy(m_main) %>%
  select(term, estimate)

m_int_1980 <- lm(avg ~ year + post1980 + year*post1980, data=annual_avg)
tidy(m_int_1980) %>%
  select(term, estimate)
```

**Overall Regression:**
$$\widehat{TempDiff} = -12.204 + 0.0107 \times year$$

**Interaction Model for Before and After 1980:**
$$\widehat{TempDiff} = -2.836 + 0.00587 year - 48.226 post1980 + 0.0243 year 
\times post1980$$

**Pre-1980:**
$$\widehat{TempDiff} = -2.836 + 0.00587 year$$

**Post-1980:**
$$\widehat {TempDiff} = -51.0624 + 0.0302 year$$

```{r 1980-lm-viz, message = FALSE, echo = FALSE}
ggplot(data=annual_avg, aes(x=year, y=avg, color=post1980)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_smooth(method="lm", se=FALSE, lwd = 2.5) +
  labs(title = "Global average land temperatures since 1900:",
       subtitle = 
         "Linear model for rates of temperature increase before/after 1980",
       x = "Year",
       y = "Average land temp.") +
  scale_color_discrete(name = "Time period", 
                       labels = c("Pre-1980", "Post-1980")) +
  scale_color_manual(values = c("#B48EAD", "#81A1C1")) +
  theme_economist_mod() +
  theme(plot.subtitle = element_text(size = 14, hjust = 0)) + 
  legend_mod()
```

```{r create-slopes-data, echo = FALSE}
set.seed(12)
slopes <- numeric(23)

for (i in 1:23) {
    fiveyrs <- annual_avg %>%
      filter(year >= i*5 + 1895 & year < i*5 + 1900)
    model <- lm(avg ~ year, data=fiveyrs) 
    model <- tidy(model) %>%
      select(estimate) %>%
      slice(2) %>%
      pull()
    slopes[i] <- model
}

slopes <- tibble(slopes)
five_year_slopes <- tibble::rowid_to_column(slopes, "year")
five_year_slopes <- five_year_slopes %>%
  mutate(yearstart = 1895 + 5 * year)
```

``` {r every-five-yrs-viz, echo = FALSE}
ggplot(data=five_year_slopes, aes(x=slopes)) +
  geom_histogram(bins=23, fill = "#A3BE8C", color = "#728562") +
  labs(title=
         "Distribution of rates of temperature increase",
       subtitle="Linear model for each five-year period",
       x = "Slope (rate of change in avg. earth surface temperature)",
       y = "Count") + 
  scale_x_continuous(breaks=seq(-0.2,0.25,0.05)) +
  theme_economist_mod() +
  theme(plot.subtitle = element_text(size = 14, hjust = 0))
```

```{r 1980-bootstrap-test, echo = FALSE}
boot_dist_slope <- numeric(5000)
for (i in 1:5000) {
  indices <- sample(1 : nrow(slopes), replace = TRUE)
  boot_diff <- slopes %>%
    slice(indices) %>%
      summarize(mean_slope = mean(slopes)) %>%
    pull()
  boot_dist_slope[i] <- boot_diff
}

boot_dist_slope <- tibble(boot_dist_slope)

boot_mean_slope <- boot_dist_slope %>%
  summarize(mean = mean(boot_dist_slope)) %>% 
  pull()

post1980slope <- 0.0302

boot_dist_slope %>% 
  mutate(extreme = ifelse(boot_dist_slope >= post1980slope, 1, 0))%>%
  summarize(p_value= mean(extreme))
```

From the above simulation-based test, we found a p-value of 0.0592. Because this
is not under $\alpha = 0.05$, we fail to reject this null hypothesis.

Here, the p-value is the probability of obtaining results in which the 
difference in average rate of change between post- and pre-1980 temperatures is 
at least as great as what was observed in the data.

In context of our question, this suggests we may have insufficient evidence of a
statistically significant increase in the average degree of temperature change 
from before 1980 to after 1980. 

After failing to reject the null hypothesis above, we then considered performing
the same test for the data before and after the year 2000. The linear model as 
shown below suggests a steeper incline than 1980.

```{r 2000-lm-viz, message = FALSE, echo = FALSE}
annual_avg <- annual_avg %>%
  mutate(post2000 = year >= 2000)

ggplot(data=annual_avg, aes(x=year, y=avg, color=post2000)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_smooth(method="lm", se=FALSE, lwd = 2.5) +
  labs(title = "Global average land temperatures since 1900:",
       subtitle = 
         "Linear model for rates of temperature increase before/after 2000",
       x = "Year",
       y = "Average land temp.") +
  scale_color_discrete(name = "Time period", 
                       labels = c("Pre-2000", "Post-2000")) +
  scale_color_manual(values = c("#B48EAD", "#81A1C1")) +
  theme_economist_mod() +
  theme(plot.subtitle = element_text(size = 14, hjust = 0)) + 
  legend_mod()
```

``` {r 2000-bootstrap-test, echo = FALSE}
post2000slope <- 0.0365

ggplot(data=boot_dist_slope, aes(x=boot_dist_slope)) +
  geom_histogram(bins=50,fill = "#8FBCBB", color = "#648382") +
  geom_vline(xintercept = post1980slope, 
              color = "#BF616A", lwd=1) +
  geom_vline(xintercept = post2000slope, 
              color = "#BF616A", lwd=1) +
  labs(title=
     "Simulated null distribution of mean \nrates of change in temperature",
     x = "Average rate of change (slope)",
     y = "Count") +
  annotate("text", 
           x = post1980slope - 0.0035, y = 235, 
           label= paste("Average observed post-1980 slope: \n", round(post1980slope, 4)),
           angle = 90, size = 3.5) +
  annotate("text", 
           x = post2000slope + 0.0035, y = 235, 
           label= paste("Average observed post-2000 slope: \n", round(post2000slope, 4)),
           angle = 90, size = 3.5) +
    theme_economist_mod()

boot_dist_slope %>% 
  mutate(extreme = ifelse(boot_dist_slope >= post2000slope, 1, 0))%>%
  summarize(p_value= mean(extreme))
```

In the hypothesis test for years after 2000, the p-value is 0.0214, so we can 
reject this null hypothesis. This p-value indicates a 0.0214 probability of 
obtaining results in which the post-2000 data points have a mean rate of change 
equal to or greater than the one observed, assuming that the rates of change 
truly are equal.

In context of our original research question, this suggests that the earth 
surface temperature changes that we examined in part 1 might be occurring at a 
quickening pace over time, with a decreasing probability of obtaining results as
or more extreme as observed (the p-values for the 1980s and 2000s tests). 
However, it should be noted that there is very limited data after 2000, so this 
p-value should not necessarily be interpreted as a practically significant 
value, and this test alone does not adequately answer our overall research 
question.

## Question 3

***Does the data provide evidence of a greater degree of net fluctuation in 
annual global mean temperatures (positive or negative variability) for years 
1980-2013 than years 1900-1980?***

```{r net-change, echo = FALSE}
ggplot(data=differences, aes(x=year, y=abs(differences))) +
  geom_bar(aes(fill=differences), stat="identity") +
  labs(title = 
    "Year-to year net differences in \nannual average temperatures since 1900",
       fill = "Net differences",
       x = "Year",
       y = "Change in annual avg. from previous year") + 
  theme_economist_mod() + 
  scale_fill_gradientn(colours = c("#EBCB8B","#D08770", "#BF616A")) +
  legend_mod()
```

### Null Hypothesis:

$H_o:\mu_{pre1980net} \leq \mu_{post1980net}$ 
  
The mean net (absolute value) year-to-year fluctuation in global average surface
temperatures after 1980 is not greater than the mean year-to-year fluctuation 
before 1980.

### Alternative Hypothesis:

$H_a:\mu_{pre1980net} > \mu_{post1980net}$

The mean net (absolute value) year-to-year fluctuation in global average surface
temperatures after 1980 is greater than the mean year-to-year fluctuation before
1980.

```{r net-change-bootstrap, echo = FALSE}
differences <- differences %>%
  mutate(net_diff = abs(differences))

post1980_net <- differences %>%
  filter(year >= 1980)  

pre1980_net <- differences %>%
  filter(year < 1980) 

set.seed(13)

net_boot_dist <- numeric(5000)
for (i in 1:5000) {
  post_boot_mean <- post1980_net %>%
    slice(sample(1:nrow(post1980_net), replace = TRUE)) %>%
    summarize(mean(net_diff)) %>%
    pull()
  pre_boot_mean <- pre1980_net %>%
    slice(sample(1:nrow(pre1980_net), replace = TRUE)) %>%
    summarize(mean(net_diff)) %>%
    pull()
  net_boot_dist[i] <- post_boot_mean - pre_boot_mean
}
net_boot_dist <- tibble(net_boot_dist)
```

```{r compare-to-20th-ci, echo = FALSE}
diff_bootstrap_interval <- net_boot_dist %>%
  summarize(lower = quantile(net_boot_dist, 0.005),
            upper = quantile(net_boot_dist, 0.995)
            )
diff_bootstrap_interval
```

Our 99% confidence interval for this hypothesis test is (-0.00639, 0.14843).
We can be 99% confident that the difference in means of average year-to-year net
fluctuation after 1980 and before 1980 lies within these bounds. 

```{r net-testing, echo = FALSE}
post1980_net_diff_mean <- post1980_net %>%
  summarize(mean(net_diff)) %>%
  pull()

pre1980_net_diff_mean <- pre1980_net %>%
  summarize(mean(net_diff)) %>%
  pull()

net_boot_mean <- net_boot_dist %>%
  summarize(mean(net_boot_dist)) %>%
  pull()

net_obs_diff <- post1980_net_diff_mean - pre1980_net_diff_mean

shifted_net <- net_boot_dist %>%
  mutate(shifted = net_boot_dist - net_boot_mean)

shifted_net %>%
  mutate(extreme = if_else(shifted >= net_obs_diff, 1, 0)) %>%
  summarize(p_val = mean(extreme)) %>%
  pull()
```

```{r net-clt-testing, echo = FALSE}
differences <- differences %>%
  mutate(post1980 = year >= 1980)

t.test(net_diff ~ post1980,
  data = differences,
  mu = 0,
  var.equal = FALSE,
  alternative = "less",
  conf.level = 0.95)
```

The p-values from both the simulation-based and CLT-based hypothesis tests 
performed here are both less than $\alpha = 0.05$, so we can reject the 
null hypothesis. In this case, the p-value represents the likelihood of 
obtaining results where the difference in mean net fluctuation of average temperatures
from year to year after 1980 versus the mean net fluctuation before 1980 is as great   
as observed or greater. In context of our question, this suggests we have sufficient 
evidence to argue that not only have temperatures increased over the 20th century,
but they have become more extreme overall from year to year for both cold and hot 
temperatures compared to earlier in the 20th century.

## Question 4

***Has North America experienced a greater change in annual average 
temperatures from the first half of the 20th century to 2010 than other 
continents?***

### Null Hypothesis:

$H_o: \mu_n \leq \mu_g$

The change in mean temperature from before 1950 to post-2010 in North America is 
not greater than the global change.  

### Alternative Hypothesis:

$H_a: \mu_n > \mu_g$

The change in mean temperature from before 1950 to post-2010 in North America is 
greater than the global change.

```{r addcontinent, echo = FALSE}
asia <- c("Afghanistan", "Bangladesh", "Burma", "China", "India","Indonesia",
          "Iran","Iraq","Japan","Pakistan","Philippines","Russia",
          "Saudi Arabia","Singapore","South Korea","Syria","Taiwan","Thailand",
          "Vietnam")
africa <- c("Angola","Congo","Côte D'Ivoire", "Egypt","Ethiopia", "Kenya",
            "Morocco","Nigeria","Senegal","Somalia","South Africa","Sudan",
            "Tanzania","Zimbabwe")
north_america <- c("Canada","Mexico","United States")
south_america <- c("Brazil","Chile","Colombia","Dominican Republic","Peru")
eur <- c("France","Germany","Italy","Spain","Ukraine","United Kingdom",
         "Turkey")
aus <- c("Australia")
  
changesbycountry <- changesbycountry %>%
  mutate(continent = case_when(
    country %in% asia ~ "asia",
    country %in% africa ~ "africa",
    country %in% north_america ~ "north_america",
    country %in% south_america ~ "south_america",
    country %in% eur ~ "eur",
    country %in% aus ~ "aus"
  ), is_europe = continent == "eur",
  is_northamerica = continent == "north_america")
```

```{r ttest-northamerica, echo = FALSE}
t.test(change ~ is_northamerica,
       data = changesbycountry,
       mu = 0,
       var.equal = FALSE,
       alternative = "less",
       conf.level = 0.95)
```

From this first hypothesis test, we fail to reject the null hypothesis that 
North America has a different change in mean temperature from 1950-2010 than the
rest of the world. The p-value was 0.2166, indicating a .2166 probability of 
obtaining results in which North America had an average difference of overall temperature
change with respect to the average difference for other continents that was as great
or greater than was observed as shown in this data set, assuming the null hypothesis.

```{r ttest-europe, echo = FALSE}
t.test(change ~ is_europe,
       data = changesbycountry,
       mu = 0,
       var.equal = FALSE,
       alternative = "less",
       conf.level = 0.95)
```

We also constructed a similar t-test for Europe, and when comparing the 
confidence intervals for the two, comparing Europe to the rest of the world 
produces a much smaller confidence interval (-0.4708, 0.3892), versus that for 
North America: (-2.2975, 1.4352). Additionally, the p-value for Europe's t-test 
was 0.4155, indicating a higher likelihood of Europe having the same true mean 
change in land temperature as the rest of the world.

```{r df-rworldmap, message = FALSE, echo = FALSE}
#For a short introduction type : 	 vignette('rworldmap')
changesbycountry$country[12] <- "Ivory Coast" #prev. fail match w/ country code
geo_country_df <- joinCountryData2Map(changesbycountry,
                                      joinCode = "NAME",
                                      nameJoinColumn = "country")
                  #specify verbose = TRUE arg. to get full country list
```

```{r viz-rworldmap, message = FALSE, echo = FALSE}
par(mai = c(0, 0, 0.2, 0), xaxs = "i", yaxs = "i")
customColorGrad <- RColorBrewer::brewer.pal(10, 'YlOrRd')
#calling mapCountryData with the parameters from RColorBrewer and geo_country_df
mapCountryData(geo_country_df, nameColumnToPlot="change",
               mapTitle= "Temperature Change 1950-2010 by Country",
               colourPalette = customColorGrad, # equiv. to "heat" colourPalette
               oceanCol="lightblue",
               missingCountryCol="white")
```

# Discussion

Throughout our analysis, we saw clear evidence that climate change exists and 
that global land and ocean temperatures are increasing to an unprecedented 
extent at an extraordinary rate. As seen in our linear regression models and 
associated scatter plots, there is meaningful evidence that the rate of 
temperature change has accelerated, especially since 1975. 

Additionally, we were able to use hypothesis testing to support our claim that 
mean annual temperatures are increasing. In our first hypothesis test, we 
concluded that there is a statistically significant increase in mean earth 
surface temperature from early-20th-century levels to the present. Furthermore, 
the mean temperature from years 1980 through 2013 is greater than the mean Earth
temperature of the entire 20th century. 

Our second hypothesis test found insufficient evidence of a statistically 
significant increase in the average degree of temperature change from before 
1980 to after 1980, but did find a statistically significant increase in average
temperature change since 2000. While the data sets have limited values for years 
after 2000, it still shows evidence that global warming trends are becoming more
severe. This could be due to numerous confounding effects, including human 
causes such as increased greenhouse gas emissions in recent years. 

Our third hypothesis test found that there is also sufficient evidence of an 
increase in global, annual year-to-year temperature change in more recent decades (post-1980) 
compared to the earlier 20th century. This result is supported by other findings from
the scientific community, such as NOAA, which establishes that the rate of annual 
temperature change has increased in recent decades: 
[climate.gov Source](https://www.climate.gov/news-features/understanding-climate/climate-change-global-temperature#:~:text=Change%20over%20time&text=According%20to%20the%20NOAA%202019,more%20than%20twice%20as%20great.)

The last hypothesis test found that both Europe and the United States of America
do not significantly differ from the mean change in temperature globally. 
However, it also confirms that the U.S. is subject to more variable temperature 
fluctuations. This could provide evidence towards the effectiveness of 
initiatives in the European Union against climate change - such as the Paris 
Climate Accord - which the USA withdrew from in 2017.

While our findings supported the existence of climate change, it is important
to also critique our data and methods. The data sets, 
`GlobalLandTemperaturesByMajorCity.csv` and `GlobalTemperatures.csv` attempt to 
comprehensively describe wide ranging global effects from 1849 to 2013. However 
there are inherent limitations in attempting this, as it fails to adequately or 
equally monitor all regions over this time period. For example, there were less 
European and South American cities included than Asian cities. This limits our 
ability to draw conclusions about trends in temperature change across Europe and
South America for the same time period. Another limitation was the lack of data 
from earlier years. The data set included many NA values in the years 1849-1900,
which is why we had to focus our analysis more specifically on the past 100 
years. We also need to express that temperature collection and recording methods
have improved over time. Thus, we have more confidence that recent 
observations are generally more accurate than observations preceding the digital 
revolution.

There were some limitations with our methods. In our difference of means 
hypothesis tests, while we found that there was not a difference between Europe 
and America’s respective average change in temperature and the global change in 
temperature, our tests could not explain the reasons why the differences did not 
exist. Another limitation of the difference of means test is that while we did 
not find enough evidence to reject the null hypothesis that there was no 
difference, it does not mean that a difference does not exist. A final 
limitation in our methods was that our hypothesis tests could not actually make 
decisions about what they were testing, they could only be used as aids in 
decision making.

If we were able to restart the project, we would have found a data set that 
already contained all of the data we wanted to analyze. While we were able to 
join the two data sets together and perform our analysis, this was tedious 
compared to just using one data set. If we used one data set, we also could have 
performed more focused analysis on the data.



