---
title: "Project - Team 10-7"
author: "The Outliers: Morgan Pruchniewski, Scott Burstein, Katie Zhou"
date: "10/6/2020"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse) 
```

Source of the Data: We accessed the data from Kaggle, but the data originally came from the Berkeley Earth Data Page.

When and How It Was Originally Collected: The Berkeley Earth Data Page collected the data from 16 pre-existing archives. It was updated to Kaggle in 2017.

The Cases: Each observation is a city or state and its respective land temperature, investigating how this changes over time. There are 239,177 cases in the major city dataset and 645,675 cases in the state dataset.

Relevant Variables: Some of the relevant variables in the dataset include dt (date and time of the recorded temperature, a discrete numeric), City, Country, AverageTemperature (the average temperature in that location during that time period), latitude, longitude.

```{r glimpse}
major_city <- read_csv("data/GlobalLandTemperaturesByMajorCity.csv")
temps <- read_csv("data/GlobalTemperatures.csv")

glimpse(major_city)
```

#Research Question:
Is there significant support for rising earth surface temperatures, and if so, how has the rate of change in temperatures fluctuated over time?

How do the changes in earth surface temperature differ geographically, and are there regions where climate change seems to have a greater effect?



