---
title: "Project - Team 10-7"
author: "The Outliers: Morgan Pruchniewski, Scott Burstein, Katie Zhou"
date: "10/6/2020"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse) 
```

Source of the Data: We downloaded the data from Kaggle, but the data was originally compiled by the Berkeley Earth Data Lab.

When and How It Was Originally Collected: The Berkeley Earth Data Page collected the data from 16 pre-existing archives. It was updated to Kaggle in 2017.

We tentatively plan to interpolate these time-series datasets with other datasets in order to draw even more interesting conclusions, potentially regarding the correlation between global warming and the prevalence of natural disasters. 

###The Cases:

(By Time and Location (for `GlobalLandTemperaturesByMajorCity`))
Each observation in `GlobalLandTemperaturesByMajorCity` is a city and its respective land temperature, coordinates, and Country, which will be used to investigate changes in climate over time. There are 239,177 rows in the major city dataset. 

(By Time (for `GlobalLandTemperaturesByMajorCity`))
Each observation in `GlobalTemperatures` is a numeric date/time value (described below) and its respective land temperature, max temperature, min temperature, ocean temperature, and relevant uncertainties to said variables, which will be used to investigate changes in climate over time. There are 239,177 rows in the global dataset. 

###Relevant Variables:

The relevant variables in the `major_city` dataset include dt (date and time of the recorded temperature, a discrete numeric), AverageTemperature, AverageTemperatureUncertainty, and geolocation data.

The relevant variables in the `global` dataset include dt (date and time of the recorded temperature, a discrete numeric), LandAverageTemperature (the average global land temperature at that time), and LandAverageTemperatureUncertainty, among many others.

```{r glimpse}
#read in datasets
major_city <- read_csv("data/GlobalLandTemperaturesByMajorCity.csv")
global <- read_csv("data/GlobalTemperatures.csv")

#summary stats for `major_city`
glimpse(major_city)
colnames(major_city)

#summary stats for `global`
glimpse(global)
colnames(global)
```

#Research Question:
Is there significant support for rising earth surface temperatures, and if so, how has the rate of change in temperatures fluctuated over time?

How do the changes in earth surface temperature differ geographically, and are there regions where climate change seems to have a greater effect?



