---
title: "Project - Team 10-7"
author: "The Outliers: Morgan Pruchniewski, Scott Burstein, Katie Zhou"
date: "10/6/2020"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse) 
library(lubridate)
library(maps)
library(sp)
```

### Source of the Data:
We downloaded the data from Kaggle, but the data was originally compiled by the 
Berkeley Earth Data Lab.

### When and How It Was Originally Collected: 
The Berkeley Earth Data Page collected the data from 16 pre-existing archives. 
It was updated to Kaggle in 2017.

We tentatively plan to interpolate these time-series datasets with other 
datasets in order to draw even more interesting conclusions, potentially 
regarding the correlation between global warming and the prevalence of natural 
disasters. 

### The Cases:

**(By Time and Location (for `GlobalLandTemperaturesByMajorCity`))**

Each observation in `GlobalLandTemperaturesByMajorCity` is a city and its 
respective land temperature, coordinates, and Country, which will be used to 
investigate changes in climate over time. There are 239,177 rows in the major 
city dataset. 

**(By Time (for `GlobalLandTemperaturesByMajorCity`))**

Each observation in `GlobalTemperatures` is a numeric date/time value 
(described below) and its respective land temperature, max temperature, 
min temperature, ocean temperature, and relevant uncertainties to said 
variables, which will be used to investigate changes in climate over time. 
There are 239,177 rows in the global dataset. 

### Relevant Variables:

The relevant variables in the `major_city` dataset include dt (date and time of 
the recorded temperature, a discrete numeric), AverageTemperature (the average 
temperature for each city, a continuous numeric), city (nominal categorical), 
country (nominal categorical), latitude (continuous numeric), and longitude 
(continuous numeric).

The relevant variables in the `global` dataset include dt (date and time of the 
recorded temperature, a discrete numeric), LandAverageTemperature (the average 
global land temperature at that time, continuous numeric), LandMaxTemperature 
(the highest recorded land temperature of that year, a continuous numeric), 
LandMinTemperature (the lowest recorded land temperature of that year, a 
continous numeric), and LandAndOceanAverageTemperature (the average of land and 
ocean temperature averages for that year, a continuous numeric).

```{r read-in-data, message = FALSE}
#read in datasets
major_city <- read_csv("data/GlobalLandTemperaturesByMajorCity.csv")
global <- read_csv("data/GlobalTemperatures.csv",
                    col_types = "Tdddddddd")
df <- left_join(major_city, global) 
```

```{r data-processing}
df$lat<-as.numeric(gsub("N|E|S|W", "",df$Latitude))*
  ifelse(grepl("S",df$Latitude),-1,1)

df$long<-as.numeric(gsub("N|E|S|W", "",df$Longitude))*
  ifelse(grepl("W",df$Longitude),-1,1)

df <- df %>%
  mutate(month = month(dt),
         year = year(dt),
         season = case_when(
           month < 3 | month == 12 ~ "winter",
           month >= 3 & month < 6 ~ "spring",
           month >=6 & month < 9 ~ "summer",
           month >= 9 & month < 12 ~ "autumn"
         )) 

df$Country <- replace(df$Country, df$Country=="Congo (Democratic Republic Of The)","Congo")
 
```

### Research Question:

**Is there significant support for rising earth surface temperatures, and if so,**
**how has the rate of change in temperatures fluctuated over time?**

$H_o$ = There is no statistical evidence of rising earth surface temperature 
since 1750. Thus, the rate of temperature change is constant (0).

$H_a$ = There is statistical evidence of rising earth surface temperature since 
1750, and the rate of annual temperature change has also changed.


```{r visualization-by-season}
ggplot(data=df, aes(
  x=dt, y=LandAverageTemperature
)) +
  geom_point(aes(color=season)) + 
  labs(title = "Global average land temperature since 1900",
       x = "Year",
       y = "Average Land Temp.")

```
Based on this initial visualization, there seems to be a high amount of uncertainty in the average land temperature in the earlier data. Going forward, we will only use the data after 1900.
Overall, across all seasons, there seems to be an increasing trend in average land temperature after roughly 1975. 

```{r filter-years}
df <- df %>%
  filter(dt >= "1900-01-01")
```



```{r temp-oceans}
ggplot(data=df, aes(x=dt, y=LandAndOceanAverageTemperature)) +
  geom_point(aes(color=season)) + 
  labs(title = "Global average land + ocean temperature since 1900",
       x = "Year",
       y = "Average Land + Ocean Temp.")
```
From these two visualizations, the average land/ocean temperature seems to increase more steeply after 1975 than the average land temperature. The values for spring and autumn in the land/ocean data are also much more similar than they are for just land.

(why are there distinct bands in the data?)


```{r annual-avg-viz}
annual_avg <- df %>%
  group_by(year) %>%
  summarize(avg = mean(LandAverageTemperature))

ggplot(data=annual_avg, aes(x=year, y=avg)) +
  geom_point(aes(color=avg), size=2) + 
  labs(title = "Global average land temperature since 1900",
       x = "Year",
       y = "Average Land Temp.")
```
When we plot the average annual land temperature, there seems to be a more extreme change in average land temp. over time than when each month is plotted individually.

```{r facet-season}
#do we need this?
ggplot(data=df, aes(x=dt, y=LandAverageTemperature)) +
  geom_point(aes(color=LandAverageTemperature)) + 
  facet_wrap(~season)
```




```{r country-avg-changes}
bycountrypre1950 <- df %>%
  filter(!is.na(AverageTemperature),
         year < 1950) %>%
  group_by(Country) %>%
  summarize(avg = mean(AverageTemperature))

bycountry2010s <- df %>%
  filter(!is.na(AverageTemperature),
         year > 2009) %>%
  group_by(Country) %>%
  summarize(avg = mean(AverageTemperature))

changesbycountry <- tibble(country=bycountry2010s$Country, 
                           change = bycountry2010s$avg - bycountrypre1950$avg) 


```

```{r visualization-by-country}
ggplot(data=changesbycountry,aes(x=reorder(country,-change),y=change)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    title="Average change from pre-1950 data to post-2010 data, by country",
    x="Country")
```

```{r highs-lows-visualization}
ggplot(data=df, aes(x = dt)) +
  geom_line(aes(y = LandMaxTemperature)) +
  geom_line(aes(y = LandMinTemperature))
```



```{r map}
world <- map_data("world")

#ggplot() +
	#geom_polygon(data = world, 
	#	aes(x = long, y = lat, group = group),
	#	fill = NA, colour = "black") + 
#	geom_point(data = bycountrypre1950,  
#		aes(x = long, y = lat, color=avg)) +
#	coord_quickmap() +  
#	theme_bw() 
```

**How do the changes in earth surface temperature differ geographically, and are**
**there regions where climate change seems to have a greater effect?**



$H_o$: The average earth surface temperature since 1980 is not greater than the average earth surface temperature over the last 120 years.

$H_a$: The average earth surface temperature data since 1980 are greater than the average earth surface temperature over the last 120 years by a statistically significant amount.


^if we reject null:
$H_o$: The average earth surface temperature since 1980 is changing at the same rate that it was changing from 1900-1930.

$H_a$: The average earth surface temperature since 1980 is changing at a higher rate rate than it was changing from 1900-1930.


```{r differences-for-loop}
differences <- c(nrow(annual_avg))

for (i in 1:nrow(annual_avg) - 1) {
  differences[i] <- annual_avg$avg[i+1] - annual_avg$avg[i]
}

```

```{r differences-create-df}
differences <- tibble(differences)
differences <- tibble::rowid_to_column(differences, "year")
differences <- differences %>%
  mutate(year = 1900 + year)

```

```{r differences-viz}

differences

ggplot(data=differences, aes(x=year, y=differences)) +
  geom_bar(stat="identity")
```

compare linear model with earlier data to linear model for later data?

global temps are greater now than they were from 1900-1930, and also fluctuating at a greater rate(?)



comparing annual avg temp to avg 20th century temp:
```{r compare-to-20th-century}
avg_20thcentury <- annual_avg %>%
  summarize(mean(avg)) %>%
  pull()

comparison_to_20th <- c(nrow(annual_avg))

for (i in 1:nrow(annual_avg) - 1) {
  comparison_to_20th[i] <- annual_avg$avg[i+1] - avg_20thcentury
}

comparison_to_20th <- tibble(comparison_to_20th)
comparison_to_20th <- tibble::rowid_to_column(comparison_to_20th, "year")
comparison_to_20th <- comparison_to_20th %>%
  mutate(year = 1900 + year)
```

```{r visualization-compare-to-avg}
ggplot(data=comparison_to_20th, aes(x=year, y=comparison_to_20th)) +
  geom_bar(stat="identity")
```
we want to test whether the average anomaly from between 1980-2010 > average anomaly from between 1900-1930?



$H_o$
Canada has experienced earth surface temperature change to the same degree as the global average.

$H_a$
Canada has experienced a greater amount of earth surface temperature change than the global average.




$H_o$
Latitude and temperature change are independent; there is no correlation.

$H_a$
Latitude and temperature change are not independent; there is a correlation between latitude and the amount of observed temperature change.







Linear model(s) for land temperature change and land+ocean temperature change?



(only if we have time)

$H_o$ = There is no variation in earth surface temperature change 
geographically.

$H_a$ = There is variation in earth surface temperature change geographically.



